### 2022-08-10 doc/src/IOMMU.tex
Replace "I/O MMU" with "IOMMU"

%=======================================================================
\chapter{I/O MMU Support for MSIs to Virtual Machines}
\label{ch:IOMMU}

@@

The existence of an I/O MMU in a system makes it possible for a guest
operating system, running in a virtual machine, to be given direct
control of an I/O device with only minimal hypervisor intervention.
A guest OS with direct control of a device will program the device with
guest physical addresses, because that is all the OS knows.
When the device then performs memory accesses using those addresses, an
I/O MMU is responsible for translating those guest physical addresses
into machine physical addresses, referencing address-translation data
structures supplied by the hypervisor.

To handle MSIs from a device controlled by a guest OS, an I/O MMU must
be able to redirect those MSIs to a guest interrupt file in an IMSIC.
Systems that do not have IMSICs with guest interrupt files do not need
to implement the facilities described in this chapter.

Because MSIs from devices are simply memory writes, they would
naturally be subject to the same address translation that an I/O MMU
applies to other memory writes.
However, the Advanced Interrupt Architecture requires that I/O MMUs
treat MSIs directed to virtual machines specially, in part to
simplify software, and in part to allow optional support for
\emph{memory-resident interrupt files}.

This chapter is concerned only with how an I/O MMU recognizes and
processes MSIs directed to virtual machines.
Most other functions and details of an I/O MMU are beyond the scope of
this standard, and must be specified elsewhere.

@@

%-----------------------------------------------------------------------
\section{Device contexts at an I/O MMU}
\label{sec:IOMMU-deviceContexts}

The following assumptions are made about the I/O MMUs in a system:
\begin{itemize}

\item
For each I/O device connected to the system through an I/O MMU,
software can configure at the I/O MMU a \emph{device context}, which
associates with the device a specific virtual address space and any
other per-device parameters the I/O MMU may support.
By giving devices each their own separate device context at an I/O MMU,
each device can be individually configured for a different software
master, usually a guest OS or the main (host) OS.
On every memory access made from a device, hardware indicates to
the I/O MMU the originating device by some form of unique device
identifier, which the I/O MMU uses to locate the appropriate device
context within data structures supplied by software.
For PCI, for example, the originating device may be identified by the
unique triple of PCI bus number, device number, and function number.

\item
An I/O MMU optionally translates the addresses of a device's memory
accesses using address-translation data structures---typically page
tables---specified by software via the corresponding device context.
The smallest granularity of address translation implemented by all
I/O MMUs is not larger than a \mbox{4-KiB} page, matching that of
standard {\RISCV} address-translation page tables.
(An I/O MMU may in fact employ page tables in the same format as the
page-based address translation defined by the {\RISCV} Privileged
Architecture, but this is not required.)

@@

First, hypervisors running under Linux or a similar OS can benefit from
separate control of MSI translations to help simplify the case when
virtual harts are migrated from one physical hart to another.
As noted in Section~\ref{sec:virtHartMigration}, when a virtual
hart's interrupt files are mapped to guest interrupt files in the
real machine, migration of the virtual hart causes the physical guest
interrupt files underlying those virtual interrupt files to change.
However, because on other systems (not {\RISCV}) the migration
of a virtual hart does not affect the mapping from guest physical
addresses to real physical addresses, the internal functions of Linux
that perform this migration are not set up to modify an I/O MMU's
address-translation tables to adjust for the changing physical
locations of {\RISCV} virtual interrupt files.
Giving a hypervisor control of a separate MSI translation table at an
I/O MMU bypasses this limitation.
The MSI page table can be modified at will by the hypervisor and/or
by the subsystem that manages interrupts without coordinating with the
many other OS components concerned with regular address translation.

@@

To support the delivery of MSIs from I/O devices directly to {\RISCV}
virtual machines without hypervisor intervention, an I/O MMU must be
able to translate the guest physical address of an MSI to the real
physical address of an IMSIC's guest interrupt file in the machine, as
illustrated in Figure~\ref{fig:IOMMU-guestIntrFiles}.
This address translation is controlled by the MSI page table configured
in the appropriate device context at the I/O MMU.
Because every interrupt file, real or virtual, occupies a naturally
aligned \mbox{4-KiB} page of address space, the required address
translation is from a virtual (guest) page address to a physical page
address, the same as supported by regular {\RISCV} page-based address
translation.

\begin{figure}[th]
\centerline{\includegraphics[scale=0.55]{IOMMU-guestIntrFiles.png}}
\caption{%
Translation of a device-sourced MSI that a guest OS intended to go to
a (virtual) IMSIC interrupt file in the OS's virtual machine.
Referencing an MSI page table supplied by the controlling hypervisor,
the I/O MMU redirects the MSI to a guest interrupt file of the real
machine.%
}
\label{fig:IOMMU-guestIntrFiles}
\end{figure}

Memory writes from a device are recognized as MSIs by the destination
address of the write.
If an I/O MMU determines that a \mbox{32-bit} write is to the location
of a (virtual) interrupt file in the relevant virtual machine, the
write is considered an MSI within the VM, else not.
The exact formula for recognizing MSIs is documented in
Section~\ref{sec:IOMMU-identIncomingMSIs}.

\begin{commentary}
Although the translation of MSIs is controlled by its own separate page
table, the fact that MSI translations are at the same page granularity
as regular {\RISCV} address translations implies that an address
translation cache within an I/O MMU requires little modification to
also cache MSI translations.
Only on a translation cache miss does the I/O MMU need to treat MSIs
significantly differently than other memory accesses from the same
device, to choose the correct translation table and to access and
interpret the table properly.
\end{commentary}

@@

An I/O MMU may optionally support memory-resident interrupt files
(MRIFs).
If implemented, the use of memory-resident interrupt files can greatly
increase the number of virtual harts that can be given direct control
of one or more physical devices in a system, assuming the rest of the
system can still handle the added load.

@@

With the use of memory-resident interrupt files, on the other hand,
the total number of virtual {\RISCV} harts able to receive device MSIs
is almost unbounded, constrained only by the amount of real physical
memory and the additional processing time needed to handle them.
As its name implies, a memory-resident interrupt file is located in
memory instead of within an IMSIC.
Figure~\ref{fig:IOMMU-MRIF} depicts how an I/O MMU can record an
incoming MSI in an MRIF.
When properly configured by a hypervisor, an I/O MMU recognizes certain
incoming MSIs as intended for a specific virtual interrupt file, and
records each such MSI by setting an interrupt-pending bit stored within
the MRIF data structure in ordinary memory.
After each MSI is recorded in an MRIF, the I/O MMU also sends a
\emph{notice MSI} to the hypervisor to inform it that the MRIF contents
may have changed.

@@

An I/O MMU can have one of these three levels of support for
memory-resident interrupt files:
\begin{tightList}
\item no memory-resident interrupt files;
\item memory-resident interrupt files without atomic update; or
\item memory-resident interrupt files with atomic update.
\end{tightList}

Memory-resident interrupt files are most efficient when the memory
system supports logical atomic memory operations (AMOs) corresponding
to {\RISCV} instructions AMOAND and AMOOR, for memory accesses made
both from harts and from the I/O MMU.
The AMOAND and AMOOR operations are required for \emph{atomic update}
of a memory-resident interrupt file.
A reduced level of support is possible without AMOs, relying solely on
basic memory reads and writes.

@@

\begin{commentary}
The interrupt-pending and interrupt-enable bits are stored interleaved
by doublewords within an MRIF to facilitate the possibility of an
I/O MMU examining the relevant enable bit to determine whether to send
a notice MSI after updating a pending bit, rather than the default
behavior of always sending a notice MSI after an update without regard
for the interrupt-enable bits.
The memory arrangement matters only when MRIFs are supported without
atomic update.
\end{commentary}

Bit~0 of the first doubleword of an MRIF stores a faux
interrupt-pending bit for nonexistent interrupt~0.
If a write from an I/O device appears to be an MSI that should be
stored in an MRIF, yet the data to write (the interrupt identity) is
zero, the I/O MMU acts as though zero were a valid interrupt identity,
setting bit~0 of the target MRIF's first doubleword and sending a
notice MSI as usual.

All MRIFs are the size to accommodate 2047 valid interrupt identities,
the maximum allowed for an IMSIC interrupt file.
If a system's actual IMSICs have interrupt files that implement
only $N$ interrupt identities, ${N < \mbox{2047}}$, then the contents
of MRIFs for identities greater than $N$ may be ignored by software.
I/O MMUs, however, treat every MRIF as though all interrupt identities
in the range 0--2047 are valid, even as software ignores invalid
identity~0 and all identities greater than~$N$.

\begin{commentary}
There is no need to specify to an I/O MMU a desired size~$N$ for an
MRIF smaller than 2047 valid interrupt identities.
The only use an I/O MMU would make of this information would be to
discard any MSIs indicating an interrupt identity greater than~$N$.
If devices are properly configured by software, such errant MSIs should
not occur;
but even if they do, it is just as effective for software to ignore
spurious interrupt identities \emph{after} they have been recorded in
an MRIF as for an I/O MMU to discard them before recording them in the
MRIF.

It is likewise unnecessary for I/O MMUs to check for and discard MSIs
indicating an invalid interrupt identity of zero.
\end{commentary}

@@

The data component of an MSI write specifies the interrupt identity to
raise in the destination interrupt file.
(Recall Section~\ref{sec:MSIEncoding}.)
This data may be in little-endian or big-endian byte order.
If an I/O MMU supports memory-resident interrupt files, it can store to
an MRIF MSIs of the same endianness that the machine's IMSICs accept.
All IMSIC interrupt files are required to accept MSIs in little-endian
byte order written to memory-mapped register \z{seteipnum\_le}
(Section~\ref{sec:IMSIC-memRegion}).
IMSIC interrupt files may also accept MSIs in big-endian byte order if
register \z{seteipnum\_be} is implemented alongside \z{seteipnum\_le}.

If the interrupt identity indicated by an MSI's data (when interpreted
in the correct byte order) is in the range 0--2047, an I/O MMU stores
the MSI to an MRIF by setting to one the interrupt-pending bit in the
MRIF for that identity.
If atomic update is supported for MRIFs, the pending bit is set using
an AMOOR operation, else it is set using a non-atomic read-modify-write
sequence.
After the interrupt-pending bit is set in the MRIF, the I/O MMU sends
the notice MSI that software has configured for the MRIF.

@@

\begin{commentary}
It is an open question whether an I/O MMU might optionally examine
the matching interrupt-enable bit within a destination MRIF to decide
whether to send a notice MSI after setting an interrupt-pending bit.
Currently, an I/O MMU is required always to send a notice MSI after
storing an MSI to an MRIF, even when the corresponding enable bit for
the interrupt identity is zero.
\end{commentary}

@@

\item
Modify all relevant translation tables at I/O MMUs so that MSIs for
this virtual interrupt file are now stored in the MRIF.
If necessary, synchronize with all I/O MMUs to ensure that no straggler
MSIs will arrive at the old IMSIC interrupt file after this step.

@@

With atomic update of MRIFs, a virtual hart may continue executing
with its interrupt file contained in an MRIF, so long as the hypervisor
emulates for the virtual hart a proper IMSIC interrupt file to hide the
underlying MRIF.
Hypervisor software can safely set and clear the interrupt-pending and
interrupt-enable bits of the MRIF using AMOOR and AMOAND operations,
even as an I/O MMU may be storing incoming MSIs into the same MRIF.

\begin{commentary}
If an I/O MMU is ever configured to examine an MRIF's interrupt-enable
bits to decide whether to send notice MSIs, then modifying those enable
bits will generally require coordination with the I/O MMU.
But so long as I/O MMUs ignore the interrupt-enable bits as is
currently assumed, the bits can be changed by software without risk.
\end{commentary}

@@

\item
Modify all relevant translation tables at I/O MMUs so that MSIs for
this virtual interrupt file are now sent to the IMSIC interrupt file.
If necessary, synchronize with all I/O MMUs to ensure that no straggler
MSIs will be stored in the MRIF after this step.

@@

First, if the I/O devices that a virtual hart controls are behind
multiple I/O~MMUs, then multiple MRIF structures are needed, one per
I/O MMU, not just a single MRIF structure.
Furthermore, in addition to locations for storing \z{eidelivery} and
\z{eithreshold}, software needs a place for a complete copy of the
interrupt file's implemented \z{eip} array, apart from the MRIFs.
While a virtual interrupt file is in memory, its interrupt-pending bits
will be split across all the MRIFs and the saved \z{eip} array.
The interrupt-enable bits will exist only in the MRIFs.

To move a virtual hart's interrupt file from an IMSIC into memory, with
one MRIF per I/O MMU:
\begin{enumerate}

@@

\item
At each I/O MMU, modify all relevant translation tables so that MSIs
for this virtual interrupt file are now stored in the individual MRIF
matched to the I/O MMU.
If necessary, synchronize with all I/O MMUs to ensure that no straggler
MSIs will arrive at the old IMSIC interrupt file after this step.

@@

While a virtual hart's interrupt file remains in memory, an interrupt
identity's true pending bit is the logical OR of its bit in all MRIFs
and its bit in the saved \z{eip} array.
All pending bits in the MRIFs start as zeros, but interrupts may become
pending there as MSIs for this virtual hart arrive at I/O MMUs and are
stored in the corresponding MRIFs.

Without atomic update of MRIFs, an interrupt-pending bit is not easily
cleared in an MRIF.
(Clearing a single pending bit in one MRIF requires that a new MRIF be
allocated and initialized and the corresponding I/O MMU reconfigured to
store MSIs into the new MRIF.)
For this reason, it may or may not be practical to have a virtual hart
execute while keeping one of its interrupt files in memory.
When an MRIF records an interrupt that should wake a virtual hart, the
simplest strategy is to always move the interrupt file back into an
IMSIC's guest interrupt file before resuming execution of the virtual
hart.

@@

\item
Modify all relevant translation tables at I/O MMUs so that MSIs for
this virtual interrupt file are now sent to the IMSIC interrupt file.
If necessary, synchronize with all I/O MMUs to ensure that no straggler
MSIs will be stored in MRIFs after this step.

@@

When an I/O device is configured directly by a guest operating system,
MSIs from the device are expected to be targeted to virtual IMSICs
within the guest OS's virtual machine, using guest physical addresses
that are inappropriate and unsafe for the real machine.
An I/O MMU must recognize certain incoming writes from such devices as
MSIs and convert them as needed for the real machine.
(Recall Figure~\ref{fig:IOMMU-guestIntrFiles}.)

@@

Whenever an I/O MMU recognizes an incoming write from a device as
an MSI by the method specified in the previous section, the MSI is
translated or converted by consulting the MSI page table configured for
the device, instead of using the regular translation data structures
that apply to all other memory accesses from the same device.

@@

The number of entries in an MSI page table is always a power of two,
specifically $\mbox{2}^{k}$ where $k$ is the number of bits that are
ones in the MSI address mask used to extract the interrupt file number
from the destination guest physical address.
If an MSI page table has 256 or fewer entries, the start of the table
is always aligned to a \mbox{4-KiB} page address in real physical
memory.
If an MSI page table has ${\mbox{2}^{k} > \mbox{256}}$
entries, the table must be naturally aligned to a
$\mbox{2}^{k}\times \mbox{16-byte}$ address boundary.
If an MSI page table is not aligned as required, all entries in the
table appear to an I/O MMU as {\unspecified}, and any address an
I/O MMU may compute and use for reading an individual MSI PTE from the
table is also {\unspecified}.

Every \mbox{16-byte} MSI PTE is interpreted as two \mbox{64-bit}
doublewords.
If an I/O MMU also references standard {\RISCV} page tables, defined by
the {\RISCV} Privileged Architecture, for regular address translation,
then the byte order for each of the two doublewords in memory,
little-endian or big-endian, should be the same as the endianness
of the regular {\RISCV} page tables configured for the same device
context.
Otherwise, the endianness of the doublewords of an MSI PTE is
implementation-defined.

Bit~0 of the first doubleword of an MSI PTE is field~V (Valid).
When V~=~0, the PTE is invalid, and all other bits of both doublewords
are ignored by an I/O MMU, making them free for software to use.

@@

When an MSI PTE has fields V~=~1, C~=~0, and W~=~1
(write-through mode), the PTE's complete format is:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l@{\quad}l]
First doubleword:  & bit 63     & C, = 0 \\
                   & bits 53:10 & PPN \\
                   & bit 2      & W, = 1 \\
                   & bit 0      & V, = 1 \\
\noalign{\medskip}
Second doubleword: & ignored \\
\end{displayLinesTable}
All other bits of the first doubleword are reserved and must be set to
zeros by software.
The second doubleword is ignored by an I/O MMU so is free for software
to use.

@@

\begin{commentary}
An I/O MMU that also employs standard {\RISCV} page tables for regular
address translation can maximize the overlap between the handling of
MSI PTEs and regular {\RISCV} leaf PTEs as follows:

For RV64, the first doubleword of an MSI PTE in write-through mode
has the same encoding as a regular {\RISCV} leaf PTE for Sv39, Sv48,
Sv39x4, or Sv48x4 page-based address translation, with PTE fields D, A,
G, U, X, and R all zeros and W~=~1.
Hence, the MSI PTE's first doubleword appears the same as a regular
PTE that grants write permission (W~=~1) but not read or execute
permissions (X~= R =~0).
This same-encoded regular PTE would translate an MSI write the same as
the actual MSI PTE, except that what would be the PTE's accessed (A),
dirty (D), and user (U) bits are all zeros.
An I/O MMU needs to treat only these three bits differently for an MSI
PTE versus a regular RV64 leaf PTE.

@@

\begin{commentary}
Typically, NPPN will be the page address of an IMSIC's interrupt file
in the real machine, and NID will be the interrupt identity to make
pending in that interrupt file to indicate that the destination MRIF
may have changed.
However, NPPN is not required to be a valid interrupt file address, and
an I/O MMU must not attempt to restrict it to only such addresses.
Any page address must be accepted for NPPN.
\end{commentary}

>>>>

%=======================================================================
\chapter{IOMMU Support for MSIs to Virtual Machines}
\label{ch:IOMMU}

@@

The existence of an \mbox{IOMMU} in a system makes it possible for a guest
operating system, running in a virtual machine, to be given direct
control of an I/O device with only minimal hypervisor intervention.
A guest OS with direct control of a device will program the device with
guest physical addresses, because that is all the OS knows.
When the device then performs memory accesses using those addresses, an
\mbox{IOMMU} is responsible for translating those guest physical addresses
into machine physical addresses, referencing address-translation data
structures supplied by the hypervisor.

To handle MSIs from a device controlled by a guest OS, an \mbox{IOMMU} must
be able to redirect those MSIs to a guest interrupt file in an IMSIC.
Systems that do not have IMSICs with guest interrupt files do not need
to implement the facilities described in this chapter.

Because MSIs from devices are simply memory writes, they would
naturally be subject to the same address translation that an \mbox{IOMMU}
applies to other memory writes.
However, the Advanced Interrupt Architecture requires that \mbox{IOMMU}s
treat MSIs directed to virtual machines specially, in part to
simplify software, and in part to allow optional support for
\emph{memory-resident interrupt files}.

This chapter is concerned only with how an \mbox{IOMMU} recognizes and
processes MSIs directed to virtual machines.
Most other functions and details of an \mbox{IOMMU} are beyond the scope of
this standard, and must be specified elsewhere.

@@

%-----------------------------------------------------------------------
\section{Device contexts at an IOMMU}
\label{sec:IOMMU-deviceContexts}

The following assumptions are made about the \mbox{IOMMU}s in a system:
\begin{itemize}

\item
For each I/O device connected to the system through an \mbox{IOMMU},
software can configure at the \mbox{IOMMU} a \emph{device context}, which
associates with the device a specific virtual address space and any
other per-device parameters the \mbox{IOMMU} may support.
By giving devices each their own separate device context at an \mbox{IOMMU},
each device can be individually configured for a different software
master, usually a guest OS or the main (host) OS.
On every memory access made from a device, hardware indicates to
the \mbox{IOMMU} the originating device by some form of unique device
identifier, which the \mbox{IOMMU} uses to locate the appropriate device
context within data structures supplied by software.
For PCI, for example, the originating device may be identified by the
unique triple of PCI bus number, device number, and function number.

\item
An \mbox{IOMMU} optionally translates the addresses of a device's memory
accesses using address-translation data structures---typically page
tables---specified by software via the corresponding device context.
The smallest granularity of address translation implemented by all
\mbox{IOMMU}s is not larger than a \mbox{4-KiB} page, matching that of
standard {\RISCV} address-translation page tables.
(An \mbox{IOMMU} may in fact employ page tables in the same format as the
page-based address translation defined by the {\RISCV} Privileged
Architecture, but this is not required.)

@@

First, hypervisors running under Linux or a similar OS can benefit from
separate control of MSI translations to help simplify the case when
virtual harts are migrated from one physical hart to another.
As noted in Section~\ref{sec:virtHartMigration}, when a virtual
hart's interrupt files are mapped to guest interrupt files in the
real machine, migration of the virtual hart causes the physical guest
interrupt files underlying those virtual interrupt files to change.
However, because on other systems (not {\RISCV}) the migration
of a virtual hart does not affect the mapping from guest physical
addresses to real physical addresses, the internal functions of Linux
that perform this migration are not set up to modify an \mbox{IOMMU}'s
address-translation tables to adjust for the changing physical
locations of {\RISCV} virtual interrupt files.
Giving a hypervisor control of a separate MSI translation table at an
\mbox{IOMMU} bypasses this limitation.
The MSI page table can be modified at will by the hypervisor and/or
by the subsystem that manages interrupts without coordinating with the
many other OS components concerned with regular address translation.

@@

To support the delivery of MSIs from I/O devices directly to {\RISCV}
virtual machines without hypervisor intervention, an \mbox{IOMMU} must be
able to translate the guest physical address of an MSI to the real
physical address of an IMSIC's guest interrupt file in the machine, as
illustrated in Figure~\ref{fig:IOMMU-guestIntrFiles}.
This address translation is controlled by the MSI page table configured
in the appropriate device context at the \mbox{IOMMU}.
Because every interrupt file, real or virtual, occupies a naturally
aligned \mbox{4-KiB} page of address space, the required address
translation is from a virtual (guest) page address to a physical page
address, the same as supported by regular {\RISCV} page-based address
translation.

\begin{figure}[th]
\centerline{\includegraphics[scale=0.55]{IOMMU-guestIntrFiles.png}}
\caption{%
Translation of a device-sourced MSI that a guest OS intended to go to
a (virtual) IMSIC interrupt file in the OS's virtual machine.
Referencing an MSI page table supplied by the controlling hypervisor,
the \mbox{IOMMU} redirects the MSI to a guest interrupt file of the real
machine.%
}
\label{fig:IOMMU-guestIntrFiles}
\end{figure}

Memory writes from a device are recognized as MSIs by the destination
address of the write.
If an \mbox{IOMMU} determines that a \mbox{32-bit} write is to the location
of a (virtual) interrupt file in the relevant virtual machine, the
write is considered an MSI within the VM, else not.
The exact formula for recognizing MSIs is documented in
Section~\ref{sec:IOMMU-identIncomingMSIs}.

\begin{commentary}
Although the translation of MSIs is controlled by its own separate page
table, the fact that MSI translations are at the same page granularity
as regular {\RISCV} address translations implies that an address
translation cache within an \mbox{IOMMU} requires little modification to
also cache MSI translations.
Only on a translation cache miss does the \mbox{IOMMU} need to treat MSIs
significantly differently than other memory accesses from the same
device, to choose the correct translation table and to access and
interpret the table properly.
\end{commentary}

@@

An \mbox{IOMMU} may optionally support memory-resident interrupt files
(MRIFs).
If implemented, the use of memory-resident interrupt files can greatly
increase the number of virtual harts that can be given direct control
of one or more physical devices in a system, assuming the rest of the
system can still handle the added load.

@@

With the use of memory-resident interrupt files, on the other hand,
the total number of virtual {\RISCV} harts able to receive device MSIs
is almost unbounded, constrained only by the amount of real physical
memory and the additional processing time needed to handle them.
As its name implies, a memory-resident interrupt file is located in
memory instead of within an IMSIC.
Figure~\ref{fig:IOMMU-MRIF} depicts how an \mbox{IOMMU} can record an
incoming MSI in an MRIF.
When properly configured by a hypervisor, an \mbox{IOMMU} recognizes certain
incoming MSIs as intended for a specific virtual interrupt file, and
records each such MSI by setting an interrupt-pending bit stored within
the MRIF data structure in ordinary memory.
After each MSI is recorded in an MRIF, the \mbox{IOMMU} also sends a
\emph{notice MSI} to the hypervisor to inform it that the MRIF contents
may have changed.

@@

An \mbox{IOMMU} can have one of these three levels of support for
memory-resident interrupt files:
\begin{tightList}
\item no memory-resident interrupt files;
\item memory-resident interrupt files without atomic update; or
\item memory-resident interrupt files with atomic update.
\end{tightList}

Memory-resident interrupt files are most efficient when the memory
system supports logical atomic memory operations (AMOs) corresponding
to {\RISCV} instructions AMOAND and AMOOR, for memory accesses made
both from harts and from the \mbox{IOMMU}.
The AMOAND and AMOOR operations are required for \emph{atomic update}
of a memory-resident interrupt file.
A reduced level of support is possible without AMOs, relying solely on
basic memory reads and writes.

@@

\begin{commentary}
The interrupt-pending and interrupt-enable bits are stored interleaved
by doublewords within an MRIF to facilitate the possibility of an
\mbox{IOMMU} examining the relevant enable bit to determine whether to send
a notice MSI after updating a pending bit, rather than the default
behavior of always sending a notice MSI after an update without regard
for the interrupt-enable bits.
The memory arrangement matters only when MRIFs are supported without
atomic update.
\end{commentary}

Bit~0 of the first doubleword of an MRIF stores a faux
interrupt-pending bit for nonexistent interrupt~0.
If a write from an I/O device appears to be an MSI that should be
stored in an MRIF, yet the data to write (the interrupt identity) is
zero, the \mbox{IOMMU} acts as though zero were a valid interrupt identity,
setting bit~0 of the target MRIF's first doubleword and sending a
notice MSI as usual.

All MRIFs are the size to accommodate 2047 valid interrupt identities,
the maximum allowed for an IMSIC interrupt file.
If a system's actual IMSICs have interrupt files that implement
only $N$ interrupt identities, ${N < \mbox{2047}}$, then the contents
of MRIFs for identities greater than $N$ may be ignored by software.
\mbox{IOMMU}s, however, treat every MRIF as though all interrupt identities
in the range 0--2047 are valid, even as software ignores invalid
identity~0 and all identities greater than~$N$.

\begin{commentary}
There is no need to specify to an \mbox{IOMMU} a desired size~$N$ for an
MRIF smaller than 2047 valid interrupt identities.
The only use an \mbox{IOMMU} would make of this information would be to
discard any MSIs indicating an interrupt identity greater than~$N$.
If devices are properly configured by software, such errant MSIs should
not occur;
but even if they do, it is just as effective for software to ignore
spurious interrupt identities \emph{after} they have been recorded in
an MRIF as for an \mbox{IOMMU} to discard them before recording them in the
MRIF.

It is likewise unnecessary for \mbox{IOMMU}s to check for and discard MSIs
indicating an invalid interrupt identity of zero.
\end{commentary}

@@

The data component of an MSI write specifies the interrupt identity to
raise in the destination interrupt file.
(Recall Section~\ref{sec:MSIEncoding}.)
This data may be in little-endian or big-endian byte order.
If an \mbox{IOMMU} supports memory-resident interrupt files, it can store to
an MRIF MSIs of the same endianness that the machine's IMSICs accept.
All IMSIC interrupt files are required to accept MSIs in little-endian
byte order written to memory-mapped register \z{seteipnum\_le}
(Section~\ref{sec:IMSIC-memRegion}).
IMSIC interrupt files may also accept MSIs in big-endian byte order if
register \z{seteipnum\_be} is implemented alongside \z{seteipnum\_le}.

If the interrupt identity indicated by an MSI's data (when interpreted
in the correct byte order) is in the range 0--2047, an \mbox{IOMMU} stores
the MSI to an MRIF by setting to one the interrupt-pending bit in the
MRIF for that identity.
If atomic update is supported for MRIFs, the pending bit is set using
an AMOOR operation, else it is set using a non-atomic read-modify-write
sequence.
After the interrupt-pending bit is set in the MRIF, the \mbox{IOMMU} sends
the notice MSI that software has configured for the MRIF.

@@

\begin{commentary}
It is an open question whether an \mbox{IOMMU} might optionally examine
the matching interrupt-enable bit within a destination MRIF to decide
whether to send a notice MSI after setting an interrupt-pending bit.
Currently, an \mbox{IOMMU} is required always to send a notice MSI after
storing an MSI to an MRIF, even when the corresponding enable bit for
the interrupt identity is zero.
\end{commentary}

@@

\item
Modify all relevant translation tables at \mbox{IOMMU}s so that MSIs for
this virtual interrupt file are now stored in the MRIF.
If necessary, synchronize with all \mbox{IOMMU}s to ensure that no straggler
MSIs will arrive at the old IMSIC interrupt file after this step.

@@

With atomic update of MRIFs, a virtual hart may continue executing
with its interrupt file contained in an MRIF, so long as the hypervisor
emulates for the virtual hart a proper IMSIC interrupt file to hide the
underlying MRIF.
Hypervisor software can safely set and clear the interrupt-pending and
interrupt-enable bits of the MRIF using AMOOR and AMOAND operations,
even as an \mbox{IOMMU} may be storing incoming MSIs into the same MRIF.

\begin{commentary}
If an \mbox{IOMMU} is ever configured to examine an MRIF's interrupt-enable
bits to decide whether to send notice MSIs, then modifying those enable
bits will generally require coordination with the \mbox{IOMMU}.
But so long as \mbox{IOMMU}s ignore the interrupt-enable bits as is
currently assumed, the bits can be changed by software without risk.
\end{commentary}

@@

\item
Modify all relevant translation tables at \mbox{IOMMU}s so that MSIs for
this virtual interrupt file are now sent to the IMSIC interrupt file.
If necessary, synchronize with all \mbox{IOMMU}s to ensure that no straggler
MSIs will be stored in the MRIF after this step.

@@

First, if the I/O devices that a virtual hart controls are behind
multiple \mbox{IOMMUs}, then multiple MRIF structures are needed, one per
\mbox{IOMMU}, not just a single MRIF structure.
Furthermore, in addition to locations for storing \z{eidelivery} and
\z{eithreshold}, software needs a place for a complete copy of the
interrupt file's implemented \z{eip} array, apart from the MRIFs.
While a virtual interrupt file is in memory, its interrupt-pending bits
will be split across all the MRIFs and the saved \z{eip} array.
The interrupt-enable bits will exist only in the MRIFs.

To move a virtual hart's interrupt file from an IMSIC into memory, with
one MRIF per \mbox{IOMMU}:
\begin{enumerate}

@@

\item
At each \mbox{IOMMU}, modify all relevant translation tables so that MSIs
for this virtual interrupt file are now stored in the individual MRIF
matched to the \mbox{IOMMU}.
If necessary, synchronize with all \mbox{IOMMU}s to ensure that no straggler
MSIs will arrive at the old IMSIC interrupt file after this step.

@@

While a virtual hart's interrupt file remains in memory, an interrupt
identity's true pending bit is the logical OR of its bit in all MRIFs
and its bit in the saved \z{eip} array.
All pending bits in the MRIFs start as zeros, but interrupts may become
pending there as MSIs for this virtual hart arrive at \mbox{IOMMU}s and are
stored in the corresponding MRIFs.

Without atomic update of MRIFs, an interrupt-pending bit is not easily
cleared in an MRIF.
(Clearing a single pending bit in one MRIF requires that a new MRIF be
allocated and initialized and the corresponding \mbox{IOMMU} reconfigured to
store MSIs into the new MRIF.)
For this reason, it may or may not be practical to have a virtual hart
execute while keeping one of its interrupt files in memory.
When an MRIF records an interrupt that should wake a virtual hart, the
simplest strategy is to always move the interrupt file back into an
IMSIC's guest interrupt file before resuming execution of the virtual
hart.

@@

\item
Modify all relevant translation tables at \mbox{IOMMU}s so that MSIs for
this virtual interrupt file are now sent to the IMSIC interrupt file.
If necessary, synchronize with all \mbox{IOMMU}s to ensure that no straggler
MSIs will be stored in MRIFs after this step.

@@

When an I/O device is configured directly by a guest operating system,
MSIs from the device are expected to be targeted to virtual IMSICs
within the guest OS's virtual machine, using guest physical addresses
that are inappropriate and unsafe for the real machine.
An \mbox{IOMMU} must recognize certain incoming writes from such devices as
MSIs and convert them as needed for the real machine.
(Recall Figure~\ref{fig:IOMMU-guestIntrFiles}.)

@@

Whenever an \mbox{IOMMU} recognizes an incoming write from a device as
an MSI by the method specified in the previous section, the MSI is
translated or converted by consulting the MSI page table configured for
the device, instead of using the regular translation data structures
that apply to all other memory accesses from the same device.

@@

The number of entries in an MSI page table is always a power of two,
specifically $\mbox{2}^{k}$ where $k$ is the number of bits that are
ones in the MSI address mask used to extract the interrupt file number
from the destination guest physical address.
If an MSI page table has 256 or fewer entries, the start of the table
is always aligned to a \mbox{4-KiB} page address in real physical
memory.
If an MSI page table has ${\mbox{2}^{k} > \mbox{256}}$
entries, the table must be naturally aligned to a
$\mbox{2}^{k}\times \mbox{16-byte}$ address boundary.
If an MSI page table is not aligned as required, all entries in the
table appear to an \mbox{IOMMU} as {\unspecified}, and any address an
\mbox{IOMMU} may compute and use for reading an individual MSI PTE from the
table is also {\unspecified}.

Every \mbox{16-byte} MSI PTE is interpreted as two \mbox{64-bit}
doublewords.
If an \mbox{IOMMU} also references standard {\RISCV} page tables, defined by
the {\RISCV} Privileged Architecture, for regular address translation,
then the byte order for each of the two doublewords in memory,
little-endian or big-endian, should be the same as the endianness
of the regular {\RISCV} page tables configured for the same device
context.
Otherwise, the endianness of the doublewords of an MSI PTE is
implementation-defined.

Bit~0 of the first doubleword of an MSI PTE is field~V (Valid).
When V~=~0, the PTE is invalid, and all other bits of both doublewords
are ignored by an \mbox{IOMMU}, making them free for software to use.

@@

When an MSI PTE has fields V~=~1, C~=~0, and W~=~1
(write-through mode), the PTE's complete format is:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l@{\quad}l]
First doubleword:  & bit 63     & C, = 0 \\
                   & bits 53:10 & PPN \\
                   & bit 2      & W, = 1 \\
                   & bit 0      & V, = 1 \\
\noalign{\medskip}
Second doubleword: & ignored \\
\end{displayLinesTable}
All other bits of the first doubleword are reserved and must be set to
zeros by software.
The second doubleword is ignored by an \mbox{IOMMU} so is free for software
to use.

@@

\begin{commentary}
An \mbox{IOMMU} that also employs standard {\RISCV} page tables for regular
address translation can maximize the overlap between the handling of
MSI PTEs and regular {\RISCV} leaf PTEs as follows:

For RV64, the first doubleword of an MSI PTE in write-through mode
has the same encoding as a regular {\RISCV} leaf PTE for Sv39, Sv48,
Sv39x4, or Sv48x4 page-based address translation, with PTE fields D, A,
G, U, X, and R all zeros and W~=~1.
Hence, the MSI PTE's first doubleword appears the same as a regular
PTE that grants write permission (W~=~1) but not read or execute
permissions (X~= R =~0).
This same-encoded regular PTE would translate an MSI write the same as
the actual MSI PTE, except that what would be the PTE's accessed (A),
dirty (D), and user (U) bits are all zeros.
An \mbox{IOMMU} needs to treat only these three bits differently for an MSI
PTE versus a regular RV64 leaf PTE.

@@

\begin{commentary}
Typically, NPPN will be the page address of an IMSIC's interrupt file
in the real machine, and NID will be the interrupt identity to make
pending in that interrupt file to indicate that the destination MRIF
may have changed.
However, NPPN is not required to be a valid interrupt file address, and
an \mbox{IOMMU} must not attempt to restrict it to only such addresses.
Any page address must be accepted for NPPN.
\end{commentary}
